Supplementary material

### 2021-11-15 analysis/paper/paper.Rmd
Set up sample_shoreline to keep info on site name and 14c correlation with artefact inventory. Need to look into details on how the topo vs. horisontal distances are measured

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, harbour and other sea-side constructions and, as is the focus of this study, Stone Age sites.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age, both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here. The results of this study can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify important societal changes, and any such variation can potentially also have implications for how archaeological investigations should be directed within the area of analysis.

# Background
Relative sea-level is so named because the parameters of interest is the elevation of the sea surface relative to land. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

@@

Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests. This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different. Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same. While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well. As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates. Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples or by pedoturbation can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report. While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.

@@

# Data
All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.

@@

horisontal and topographic distance between site and shore-line was measured.  
This process was repeated 1000 times for each phase for each site. The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3. At 1000 simulation runs the analysis of each site ranged from  hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.

@@

<div id="refs"></div>

>>>>

# Introduction

The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, harbour constructions and, as is the focus of this study, Stone Age sites.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated various and multiple sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012; @lewis2021], a similar approach is adopted here.

# Background

Relative sea-level is so named because the parameters of interest is the elevation of the surface of the sea relative to land. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

@@

Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests. This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different. Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same. While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well. As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates. Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report. While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.

@@

# Data

All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.

@@

horisontal and topographic distance between site and shore-line was measured.\
This process was repeated 1000 times for each phase for each site. The choice of 1000 iterations follows an evaluation of convergence on median distances using the example site Hovland 3. At 1000 simulation runs the analysis of each site ranged from hours, depending on the distance to the simulated shorelines and consequently the necessary size of the window of analysis.

@@

::: {#refs}
:::

### 2021-11-15 analysis/supplementary-materials/supplementary.Rmd
Set up sample_shoreline to keep info on site name and 14c correlation with artefact inventory. Need to look into details on how the topo vs. horisontal distances are measured

This document includes some notes and shows the simulation results for each individual site analysed in the study. The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site if there are more than a single date (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age). The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal. Each phase is assumed to be independent of other phases. A red outline indicates that the date(s) were not seen as related to the lithic inventory in the original report (see also main text). The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e. highways and railways impacting the adjusted sea-level have been removed and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R). The third plot on the first row shows the site location within the study area relative to the isobases of the displacement curves in use.

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table
```{r anvik, include=FALSE}

```
Anvik is situated by the highway and a smaller road. Neither appear to significantly impact the simulation results. The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c. 8500-8250 BCE).

@@

Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated. While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures. As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.

>>>>

This document includes some notes and shows the simulation results for each individual site analysed in the study. The first plot on the first row for each site shows all calibrated radiocarbon dates from the Stone Age associated with the site (see analysis/data/raw_data/radiocarbon.csv for all dates, including those falling outside of the Stone Age). The fill colour of the radiocarbon dates indicates whether they are interpreted as belonging to the same phase, and each sum indicates that the dates for a phase have been modelled using the Boundary function and then summed using OxCal. Each phase is assumed to be independent of other phases. A red outline indicates that the date(s) were not seen as related to the artefact inventory in the original report (see also main text). The second plot on the first row displays the site location in the present day landscape on the edited DTM (i.e. highways and railways impacting the adjusted sea-level have been removed, and elevation values are interpolated from the surroundings; see main text and analysis/script/03script.R). The third plot on the first row shows the site location within the study area relative to the isobases of the shoreline displacement curves in use.

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table.

```{r anvik, include=FALSE}

Anvik is situated by the highway and a smaller road. Neither appear to significantly impact the simulation results. The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c. 8500-8250 BCE).

```{r gunnarsrod4, include=FALSE}

@@

Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated. While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures. As the simulation results compared to the site map in the report [@reitan2014d], in combination indicate that the site would not have been situated any significant distance from the shoreline, the values have been set to zero for the final aggregative analysis. A note can also be made on the date that is seen as unrelated to the artefact inventory. This was done on charcoal from a pit with ceramics which had a far younger date. The date was therefore interpreted as the result of contamination of some kind [@reitan2014d].
